Multi-level modeling with Openflexo/\FML

### 2021-10-14 responseLetter.tex

\Editor{}

\begin{response}{the reviewers have made suggestions for improving the presentation but also raised a number of issues requiring clarification. An overarching theme is the request for a more explicit description of how the multi-level modelling emulation is achieved and what its general advantages and limitations are, including a clarification of what the implications would be of supporting more "levels" than the challenge required.}
%We have been more explicit to explain how we use the FML language at the end of the introduction, and in various figures. TODO.

We have improved the presentation of both, Openflexo/FML and our solution to the challenge so that: 1) the techniques we use to \emph{achieve} multi-level capabilities are more explicit; and, 2) the advantages and disadvantages of using them in order to fulfil the challenge requirements become clearer. Notably, we have added many details in the final paragraphs of the introduction section, improved the Section 2, devoted to the description of OpenFlexo, refactored figures 2, 4, 5, 6, 7 and 9 and, added figures 3 and 10.
\end{response}

\begin{response}{We very much welcome a "conventional" response to the challenge but would like to see the respective reviewer questions addressed in a revised version which is explicit as possible about the advantages and the disadvantages of the emulation. You may find the references listed below useful to keep explanations concise by referencing related work (note that reviewer 3’s "Type-Instance pattern” is also known as the “Type Object” pattern). You should also respond to other concerns reviewers raise, either by making changes to the submission or in the form of an authors' response.}

@@

Done.
\end{response}

\pagebreak

\Reviewer{\#1}
\begin{response}{In section 2 Technology the authors describe the foundations of the Openflexo approach. Here it would greatly help the reader to see an example how the approach works for traditional modeling languages. For example, how would you realize a small subset of BPMN with the approach? This could briefly be illustrated so as to give the reader an immediate understanding. Especially, the 'metamodel'-agnostic approach seems very interesting but I would like to see how that is used for traditional modeling approaches.}

The Figure \ref{fig:BPMNSubsetExample} (above) shows how to realize a small subset of BPMN with Openflexo/FML (assuming a metamodel of the full specification of BPMN exists). Concretely, we show  how a virtual model (SimplifiedBPMN) can filter an existing model (BPMN) and be instantiated. Note that connection with external sources is not required for our solution to the challenge and is only represented in order to illustrate the capabilities of Openflexo/FML. Therefore, in the article we choose not to give this level of details and focus on the left part of the figure.
% Fabien OK pour moi

@@

\textcolor{orange}{We agree with the reviewer. The previous version of the figure depicting the main concepts of the FML language (Figure 2) was too succinct. We have fully refactored it so that now important concepts such as \emph{Type} are shown. We have correspondingly improved the text explaining the figure too. We hope this changes make understanding the core concepts of the FML language easier.}

%We've changed figure 2. It shows \texttt{Type} and the main concepts used by FML. %Explanation of figure has been improved.
\end{response}

\begin{response}{In Figure 5, the base metamodel for representing processes is shown - however it is not linked to the concepts you introduced before. Would it be possible to maybe graphically highlight how it relates to the metamodel in Figure 2? Besides, as multi-level modeling is much different to traditional modeling, you may want to consider using a different notation rather than UML class diagrams - as these are commonly known for traditional modeling approaches. When reading the paper, I actually marked the references to types in yellow in my copy - ie from ProcessType to TaskType and from ActorType to ActorType (on the attribute levels) - this helped me to see more clearly where you have references from attributes to types - but that's just my way of doing it.}

\textcolor{orange}{The only two concepts from Figured 2 represented elsewhere in the paper are the \textsf{VirtualModel} and \textsf{FlexoConcept} elements, which are represented in blue in figures 3, 5, 6, 7 and 9. Then, we use the brown color to represent the instantiation of these concepts. We hope this makes the connection between the Figure 2 and the other representations in the paper clearer. Regarding the notation, it is indeed similar to UML but it has different semantics as explained in the paper.}

@@

\textcolor{red}{Propale de Fabien pour complèter la réponse ici. We have added a discussion (où?) on the fact that our approach reifies the levels using "generic" models and rely on the process of construction to implement the levels of the challenge. We hope, it makes it easier to see that it is a contribution to multi-level domain.}

%\todo{Answering this is going to be tricky!}
\end{response}

\begin{response}{The paper gives us a short introduction (somewhat overlapping with the abstract, even at the level of sentences) and then the technology is detailed. The section is a bit hard to understand at first, the difference and the relationship between federated and virtual models could be explained in more detail especially considered that the challenge solution uses virtual models only. It is not clear, how federation works in this case.}

\textcolor{orange}{We have refactored the Technology section in order to make it, hopefully, easier to follow. Apart from improvements in the text, we have notably improved Figure 2 so that now a lot more of details are visible and contributed a new figure (Figure 3) showing how the FML concepts are used in a classical modeling scenario. Note however that we use model federation only for the tooling part of our answer to the challenge. In practice, we rely solely in the reification of models through the \textsf{VirtualModel} concept in order to emulate multi-levels as shown in Figure 4.}

@@

\begin{response}{In Section 4.1.2 a behavior called newProcess is defined. Is it called automatically (by the framework) when creating a new process, or should the users call it? Is this a convention, or are such methods configured somehow to replace the usual constructors? According to the text, "same pattern applies" in several cases, but these patterns could be also included, because it is not completely clear to what part of the behavior "same" refers. In same cases, the diagram uses "/type", while in the code it is "type" -- why? The specialization of Actors is encoded by FML, wouldn’t be easier to use the built-in specialization instead?}

{\color{teal}The behavior \texttt{newProcess} is defined in the \texttt{ProcessType} concept following a \emph{factory} design pattern. Rather than being created directly a process instance is therefore created from its type. This realizes the ontological instantiation. }
\textcolor{red}{
%- newProcess est un vrai behaviour défini dans ProcessType (pattern factory) et permet de spécifier l'instantiation d'un Process à partir d'un ProcessType (c'est la partie réalisation de l'instantiation ontologique)
- difficile de répondre sur le "same pattern applies" /type raffine type défini à l'étage plus haut (fd ne comprends pas la question)
- pas compris la question sur la spécialisation des Actors (fd ne comprends pas la question)
}

\end{response}

\begin{response}{In Section 4.2 it is mentioned, where the solution uses linguistic instantiation, but it is not clear by which component it is handled and what rules are followed when instantiating the concepts. In Section 4.3, tooling is elaborated. I suggest removing redundant information about the two editors and add more details on how the concrete syntax (e.g. the texts in the box of Fig. 9.) is defined.}
We have improved the description of the relationship between FML and our metamodels (use of colors in many figures).
The tooling uses model federation, but the description of our answer relies only on \textsf{VirtualModel} and \textsf{FlexoConcept} of FML (figure 2). The tooling is suggested in figure (new) 9. The tooling is not detailed, but the VirtualModel ProcessEditor presented in figure (new) 9 defines a lot of interactions, where texts of (old) figure 9 are found.

The business logic of the diagram tooling in encoded in FML, and more particulary for this editor in \texttt{ProcessEditor} virtual model presented figure 9. Structure of diagramming, interactions and presentation layer (eg texts) are defined in various concepts defined in the virtual model \texttt{ProcessEditor}.

@@

\pagebreak

\Reviewer{\#3}

\begin{response}{(1) The Type-Instance pattern is key to the authors approach to the challenge. However, the benefits of this pattern are not explored in the context of MLM. For example, the authors do not really explain approaches such as Clabjects or potency levels as they compare to the Type-Instance pattern (or the challenge). Why it the pattern better? Is it just a different approach? What can it do that other approaches cannot? Perhaps the benefit is that the Type-Instance pattern can be implemented in most modelling and implementation technologies whereas other approaches need special facilities? Finally, in terms of the pattern, why should anyone care - why would should they use the pattern?}

\textcolor{orange}{The reviewer is absolutely right. We decided to use the type-object pattern as it was easier to integrate as an ad-hoc mechanism in our solution to the challenge than other multi-level enabling mechanisms. In that sense we would like to remark that this, together with the \emph{virtual model} notion of the Openflexo/FML framework sufficed to meet all the requirements of the challenge. Note however that we intend to explore the integration of other mechanisms (such as clabjets or potency) as future work.}

### 2021-10-14 10-conclusions.tex
typo a FML, UML + trailing space

We fulfill all the challenge requirements and we propose a tooling that
demonstrates the usability of our solution. We developed an ad hoc solution,
including models and metamodels, thanks to the flexibility provided by the
metametamodeling infrastructure offered by Openflexo. This infrastructure helped us to overcome the limitations of the strict modeling paradigm mentioned in Section \ref{sec:introduction}, this is, the lack of support for different forms of classifications and for the duality type/object. In that sense, our solution seamlessly integrates the type-object pattern which allows us to dynamically create different \emph{instances} of model elements (e.g., \emph{TaskType}, among others) and use them for typing other model elements. This, together with the FML core support for (level agnostic) classical linguistic instantiation, permitted us to merge linguistic and ontological instantiation in order to provide a satisfactory solution to the Multi-Level Process challenge.

>>>>

We fulfill all the challenge requirements and we propose a tooling that
demonstrates the usability of our solution. We developed an ad hoc solution,
including models and metamodels, thanks to the flexibility provided by the
metametamodeling infrastructure offered by Openflexo. This infrastructure helped us to overcome the limitations of the strict modeling paradigm mentioned in Section \ref{sec:introduction}, this is, the lack of support for different forms of classifications and for the duality type/object. In that sense, our solution seamlessly integrates the type-object pattern which allows us to dynamically create different \emph{instances} of model elements (e.g., \emph{TaskType}, among others) and use them for typing other model elements. This, together with the \FML core support for (level agnostic) classical linguistic instantiation, permitted us to merge linguistic and ontological instantiation in order to provide a satisfactory solution to the Multi-Level Process challenge.

### 2021-10-14 4-technology.tex
typo a FML, UML + trailing space

A UML like representation of a simplified metamodel of \FML is provided in figure~\ref{fig:mm}. \FML is language designed to define models as virtual models. A \emph{virtual model} is composed of a set of \emph{concepts}, while itself being a \emph{concept}. Hence, \emph{virtual models} are structuring units forming architectures while \emph{concepts} are the core entities. A concept has a set of \emph{properties} and \emph{behaviors}. \emph{Properties} can be either simple variables, roles (pointers to a modeling element in a \emph{virtual model} or an external technology-specific model), or properties bound to complex control graphs. \FML is designed to define not only the structure of virtual models but also the collection of actions that can be performed on them. These actions are called \emph{behaviors} and rely on behavioral primitives called \emph{EditionAction}. These \emph{behaviors} can either be called or triggered by events. The reactive behaviors are useful when a federated model evolution needs to trigger a computation. Note that we do not exploit this possibility in the challenge.

@@

The tool support for model federation framework,
Openflexo\footnote{\url{https://github.com/openflexo-team}}, is
developed as an open source initiative. This tool offers a \FML
execution engine with an interactive virtual model design environment.

>>>>

A \UML like representation of a simplified metamodel of \FML is provided in figure~\ref{fig:mm}. \FML is language designed to define models as virtual models. A \emph{virtual model} is composed of a set of \emph{concepts}, while itself being a \emph{concept}. Hence, \emph{virtual models} are structuring units forming architectures while \emph{concepts} are the core entities. A concept has a set of \emph{properties} and \emph{behaviors}. \emph{Properties} can be either simple variables, roles (pointers to a modeling element in a \emph{virtual model} or an external technology-specific model), or properties bound to complex control graphs. \FML is designed to define not only the structure of virtual models but also the collection of actions that can be performed on them. These actions are called \emph{behaviors} and rely on behavioral primitives called \emph{EditionAction}. These \emph{behaviors} can either be called or triggered by events. The reactive behaviors are useful when a federated model evolution needs to trigger a computation. Note that we do not exploit this possibility in the challenge.

@@

The tool support for model federation framework,
Openflexo\footnote{\url{https://github.com/openflexo-team}}, is
developed as an open source initiative. This tool offers an \FML
execution engine with an interactive virtual model design environment.

### 2021-10-14 6-modelpresentation.tex
typo a FML, UML + trailing space

% Expliquer que le use case XSure ne nécessitait pas l'extension de metamodèle requise pour le usecase Acme Software Developement Process

>>>>

Note that in all following figures, blue is used to represent \FML virtual models and flexo concepts (\textsf{VirtualModel} and \textsf{FlexoConcept} in figure \ref{fig:mm}) , see figures \ref{fig:BPMNSubsetExample}, \ref{fig:MultilevelArchitecture}, \ref{fig:LinguisticAndOntologicInstantiation}, \ref{fig:BaseMetamodel}, \ref{fig:AcmeFullArchitecture},  \ref{fig:ToolingArchitecture}, while brown is used for instances of \textsf{FlexoConcept}, see figures \ref{fig:BPMNSubsetExample}, \ref{fig:MultilevelArchitecture}, \ref{fig:LinguisticAndOntologicInstantiation}, \ref{fig:AcmeFullArchitecture},  \ref{fig:ToolingArchitecture}.
% 3, 4, 5, 6, 7, 8, 9
% Expliquer que le use case XSure ne nécessitait pas l'extension de metamodèle requise pour le usecase Acme Software Developement Process
